---
title: "lab_02"
author: "Derek Willis, adapted by Aidan Hughes"
date: "2024-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since July 21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters.

Which counties have seen the most? Take a look at county population figures here and compare that list with your data findings in your answer: https://worldpopulationreview.com/us-counties/maryland. 
Note: here will use group_by and summarize

**A1**

17076 voters

In descending order
Montgomery County, Baltimore City, Baltimore County, Prince George's County, Anne Arundel County

```{r}
# Combining filters (ref Task 12 pre-lab)
# fyi try putting thing filtering for in quotes instead of filtering by dates
# make new dataframe/dataset/variable (variable as in that shows in Environment window under Data, not column) called new_md_voters_2024
# new_md_voters_2024 holds data from md_voters_2024...
new_md_voters_2024 <- md_voters_2024 |>
  # where both state data and county date are July 21
  filter(STATE_REGISTRATION_DATE == COUNTY_REGISTRATION_DATE) |>
  filter(COUNTY_REGISTRATION_DATE > "2024-07-21")
  # then see the number of rows in this new dataframe
  nrow(new_md_voters_2024)
  # which will give the number of voters who are new registrants since July 21
```

```{r}
new_md_voters_2024 |>
  group_by(COUNTY) |>
  summarize(total_voters = n()) |>
  arrange(desc(total_voters))
```


**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

Note: Google what function use, turn weeks into numbers ; e.g. convert Jan 3 to week 1. Google "How to turn date into week number" and use one of the functions you will see are options

Note fyi not related to this topic specifically:
When you need to rewrite code in a code block, instead of retyping it and trying to run only that code again, go up to the first time you read the file needed (e.g. here, md_voters_2024 <- read_rds("data/lab_02.rds")); run that code block and every one after it until you get to the code you want to edit. Edit it and run it, and thus no error message.

Note: July 21 is week . Aug 7 is week 

**A2**
Running out of time before deadline. I would find what number week the two above dates are ^ and then I would prob change what I did earlier in the second codeblock below to asc (so the default of arrange) of COUNTY_REGISTRATION_WEEK to answer this


```{r}
md_voters_2024 <- md_voters_2024 |>
  mutate(COUNTY_REGISTRATION_WEEK=week(COUNTY_REGISTRATION_DATE))
```

```{r}
md_voters_2024 |>
  group_by(COUNTY_REGISTRATION_WEEK) |>
  summarize(total_voters = n()) |>
  arrange(desc(total_voters))
```


**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?
**A3** Okay so this is where I would use what I typed earlier in above codeblock. But with month. I know week 24 above, but I would look at month instead here.

```{r}

```

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are.

Which regions and parties account for the most new voters?

**A4** 
use mutate to add region column... 
```{r}
```


**Q5** Filter md_voters_2024 to only include voters from the Capital region or the Western region. Save the data in a new variable/data set named capital_or_western.

How many voters are in the new data set?

```{r}

```


**Q6** Finally, filter md_voters_2024 to only include voters who meet BOTH of these conditions:

1) they are from the Eastern Shore region or Western region
2) their party affiliation is DEM and their Gender is F

Hint: you can accomplish this with two filters

Save the data in a new variable/data set named final_filter.

How many voters are in the new data set?

```{r}

```




**BONUS QUESTION** What's an additional journalistic question you could ask of this data? Show me the code you'd use to do it below, or at least describe the logic step-by-step in English if you're not sure how to write the code for it yet.


**A** 

```{r}

```

